---
title: "AE03 - Exploring gerrymandring"
author: Alexander Stevens
date: "2026-09-10"
format: html
execute: 
  warning: false
  message: false
---

## Getting started

Load the `tidyverse`, `usdata`, and `sjPlot` packages, and get the `gerrymander` data out of the `usdata` package. View the data using `glimpse`.

```{r}
library(tidyverse)
library(usdata)
library(sjPlot)
library(gtsummary)

gerrymander <- usdata::gerrymander
```

## 1. Congressional districts per state

Which state has the most congressional districts? The least? How many congressional districts are there in this state? Use functions such as `table`, `tbl_summary` or `count`.

- California has the most congressional districts at 53 while Alaska, Delaware, Montana, North Dakota, South Dakota, Vermont, and Wyoming all have 1 district for congress.

```{r}
gerrymander %>%
  select(state, district) %>%
  count(state) %>%
  arrange(n)
```

## 2. Districts at the tails

Make side-by-side box plots of percent of vote received by Trump in the 2016 Presidential Election by prevalence of gerrymandering. Identify any Congressional Districts that are potential outliers — points that fall well outside the whiskers of their box, meaning that district's Trump vote share was unusually high or low compared to other districts with the same gerrymandering prevalence. What information can you share about them?

- There are only two points that fall outside the whiskers of the box plots, and those points are districts NY-13 and NY-15. These districts voted overwhelmingly against Trump. In both districts Trump was only voted for by about 5% of the voters. These districts were both in New York and were represented by democratic congressmen.

```{r}
ggplot(gerrymander, aes(x = gerry, y = trump16)) +
  geom_boxplot() +
  labs(title = "Gerrymandering prevalence based on percentage of Trump votes",
         x = "Gerrymandering prevalence",
         y = "Trump yes vote percentage") 

gerrymander %>%
  filter(gerry == "low", trump16 < 10)

```

## 3. Gerrymandering and flipping

This relationship can be looked at from two directions — does gerrymandering predict a 2018 flip, and does a 2018 flip predict gerrymandering? Support **both** with a visualization and summary statistics. The plot answers "what does the mix look like within each group?"; the table gives the exact proportions the plot is showing. Write 1-2 bullet points about what you noticed from each plot. Use specific numbers in your answer.

- I noticed seats that didn't flip to democrat in the 2018 election were much more likely to be highly gerrymandered. For example, seats with a low gerrymandering flipped to democrat at about 26% while seats with a high gerrymandering amount only flipped at a rate of about 9%

- Medium amounts of gerrymandering also increase republican likelihood of keeping a district, suggesting a positive linear relationship between gerrymandering and likelihood of republicans keeping their seat.

**Direction 1:** Is a Congressional District more likely to have high prevalence of gerrymandering if a Democrat was able to flip the seat in the 2018 election? To compare the breakdown of gerrymandering prevalence within each flipped group, make sure `flip18` is on the x axis and that your percents add up to 100% within each category.

- Seats that didn't flip to democrats were likelier to be gerrymandered than seats that did flip to democrat in the 2018 election. 13% of seats that flipped were highly gerrymandered while 26% of seats that didn't flip were highly gerrymandered.

```{r}
gerrymander %>%
  filter(party16 == "R") %>%
  ggplot(aes(x = flip18, fill = gerry)) +
  geom_bar(position = "fill") +
  labs(title = "Republican districts in 2016 that flipped or didn't flip based on gerrymandering amount",
         x = "Did the district flip in 2018? (0 is no and 1 is yes)",
         y = "Gerrymandering percentage",
         fill = "Gerrymandering Level"
       ) 

gerrymander %>%
  filter(party16 == "R") %>%
  select(gerry, flip18) %>%
  group_by(flip18, gerry) %>%
  summarize(count = n()) %>%
  mutate(percent = count / sum(count) * 100)
```


**Direction 2:** Is a Congressional District more likely to be flipped to a Democratic seat if it has high prevalence of gerrymandering, or low prevalence? Same two variables, just flipped — so now `gerry` is on the x-axis. Again, make sure your percents add up to 100% within each category.

- Seats with a low gerrymandering amount have the highest chance to flip, at 26%. Seats with a high gerrymandering amount only have a 9% chance to flip. Seats with a medium gerrymandering amount have a 16% chance to flip, so this data shows that the higher the gerrymandering amount, the less likely a seat was to flip to democrat in the 2018 election.

```{r}
gerrymander %>%
  filter(party16 == "R") %>%
  ggplot(aes(x = gerry, fill = factor(flip18))) + # bar chart needs categorical variable so use factor on numerical variable to convert to categorical
  geom_bar(position = "fill") +
  labs(title = "Gerrymandering levels and district flipping to democrat in 2018",
        x = "Gerrymandering Level",
        y = "Flip percentage",
        fill = "Flipped to democrat? (0 is false 1 is true)"
      ) 

gerrymander %>%
  filter(party16 == "R") %>%
  select(gerry, flip18) %>%
  group_by(gerry, flip18) %>%
  summarize(count = n()) %>%
  mutate(percent = count / sum(count) * 100)
```


## 4. Gerrymandering and presidential margin

Using `mutate`, create a new variable, `margin16`, measuring how many percentage points Clinton won or lost by in each district (`clinton16 - trump16`). Does gerrymandering prevalence relate to how lopsided the 2016 presidential race was in that district? Report the min, mean, and max margin by `gerry` level, then visualize the full distribution with a violin + boxplot combo. Describe the shape, center, and spread of each group.

- According to the gerrymandering data, the average margin for districts with low gerrymandering is about 6.3%, while districts with medium gerrymandering have a margin of about 3.3% and districts with high gerrymandering have a margin of about 1.3%. This indicates the more the gerrymandered the district is, the smaller the vote margin between Trump and Hillary is. The graphs tend to center around 0 and slowly become less populous as we move away from 0, with the exceptions of a few curves becoming more populous near the extremes. The low gerrymandering group has a smaller spread than the others with the exception of a few outlier districts that voted overwhelmingly for Clinton. The high prevalence of gerrymandering group had the largest spread in comparison to the others, although the difference is not that significant, as the box is only slightly bigger.

```{r}
gerrymander %>%
  mutate(margin16 = clinton16 - trump16) %>%
  ggplot(aes(x = gerry, y = margin16)) +
  geom_boxplot() +
  labs(title = "Trump vs Clinton vote margin based on gerrymandering",
         x = "Gerrymandering prevalence",
         y = "Trump vs Clinton vote margin") 

gerrymander %>%
  mutate(margin16 = clinton16 - trump16) %>%
  ggplot(aes(x = gerry, y = margin16)) +
  geom_violin() +
  labs(title = "Trump vs Clinton vote margin based on gerrymandering",
       x = "Gerrymandering prevalence",
       y = "Trump vs Clinton vote margin") 

gerrymander %>%
  mutate(margin16 = clinton16 - trump16) %>%
  select(margin16, gerry) %>%
  group_by(gerry) %>%
  summarize(
      min_margin  = min(margin16),
      mean_margin = mean(margin16),
      max_margin  = max(margin16)
    )  
```

